---
title: "milorGWAS package"
subtitle: 'Version 0.2'
author: "Hervé Perdry, Jacqueline Milet"
version: 0.2
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{milorGWAS package}
  %\VignetteDepends{milorGWAS}
  %\VignettePackage{milorGWAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r echo = FALSE}
options(width = 120)
```

<style type="text/css">
h1.title {
  font-size: 200%;
}

h1 {
  font-size: 160%;
}

h2 {
  font-size: 130%;
}

pre {
  padding: 0px;
  margin: 0 0 0.5em 0;
}

div.sourceCode {
  margin: 0px;
}

</style>

# Introduction

This package is for Genome Wide Association Studies using a logistic mixed model, using
fast approximate methods as described in Milet et al 2020 (to appear). 
A similar method was described in Zhou (...).

Additionnally, it can draw QQ-plots with a separation of SNPs in strata, as defined 
by Chen (...) and extended in Milet et al.

This package relies on the package `gaston`, which we will use in the following.
We are going to illustrate below the functions exported by `milorGWAS`: `association.test.logistic`, `qqplot.pvalues`
(which replaces and extends the function `gaston::qqplot.pvalues`), and `SNP.category`.

# Running `association.test.logistic`

## Building a small data set
In the following code, after loading the package, we use an example from `gaston`
to simulate a binary phenotype with a random component $\omega \sim N(0, \tau K)$,
in the model
$$ \text{logit} P(Y = 1) = X \beta + \omega.  $$ 


The following lines creates a small genotype matrix `x` containing data from 1000 genomes, for 503 europeans and 733 SNPs on
a region including the TTN gene (`?TTN` for details). The simulation of the phenotype can be
done as follows.

```{r example11, echo=TRUE, message=FALSE}
library(milorGWAS)
x <- as.bed.matrix(TTN.gen, TTN.fam, TTN.bim)
x
options(gaston.verbose = FALSE)
```

Let's simulate a random phenotype.
```{r example12, echo=TRUE, message=FALSE}
set.seed(1)
## some covariates : an intercept, and a uniformly distributed covariate.
X <- cbind(1, runif(nrow(x)))
## A random GRM
ran <- random.pm(nrow(x))
## random effects (with variance tau = 1)
omega <- lmm.simu(1, 0, eigenK=ran$eigen)$omega
## linear term of the model
L <- X %*% c(0.1,-0.2) + omega
## vector of probabilities p = expit(L)
p <- 1/(1+exp( -L ))
## vector of binary phenotypes
y <- rbinom(length(p), 1, p)
```

## Logistic mixed model with Gaston

The package `gaston` can analyze these data with the Penalized Quasi Likelihood method (`a0` below), which is too
computationnaly heavy to scale to a GWAS, or with a score test, similar to GMMAT (Chen et al) (`a1` below),
which is fast but does not estimate $\beta$'s for each SNP.

```{r assoGaston, cache=TRUE}
a.pql <- association.test(x, y, X, K = ran$K, method = "lmm", response = "bin", test = "wald")
a.gmm <- association.test(x, y, X, K = ran$K, method = "lmm", response = "bin", test = "score")
```

Here are the results for the first 6 SNPs of the data set, which give similar $p$-values.

```{r}
head(a.pql)
head(a.gmm)
```

## Logistic mixed model with milorGWAS

MilorGWAS proposes two more methods, named `offset` and `amle` (Milet et al for details). 
Again we show the results for the six first SNPs.

```{r assoMilor, cache=TRUE}
a.ofst <- association.test.logistic(x, y, X, K = ran$K, algorithm = "offset")
a.amle <- association.test.logistic(x, y, X, K = ran$K, algorithm = "amle")
head(a.ofst)
head(a.amle)
```

The $p$-value from the `amle` method is always the same as the $p$-value from the score test (GMMAT). 

## Comparing the results

We can compare the $\beta$'s obtained by the `amle` or `offset` algorithm with the values obtained with the PQL:
```{r plotbeta, fig.height = 4, fig.width = 7, dev = 'png'}
par(mfrow=c(1,2), cex = 0.9)
plot(a.amle$beta, a.pql$beta, xlab = "AMLE", ylab = "PQL"); abline(0,1,col=4)
plot(a.ofst$beta, a.pql$beta, xlab = "Offset", ylab = "PQL"); abline(0,1,col=4)
```

A similar comparison for the $p$-values (remember that `amle` gives the same $p$-values than the score test).
```{r plotp, fig.height = 4, fig.width = 7, dev = 'png'}
par(mfrow=c(1,2), cex = 0.9)
plot(a.amle$p, a.pql$p, log = "xy", xlab = "AMLE/score", ylab = "PQL"); abline(0,1,col=4)
plot(a.ofst$p, a.pql$p, log = "xy", xlab = "Offset", ylab = "PQL"); abline(0,1,col=4)
```

# Stratified qq-plots

Here we will use a simulated structured population with a large number of SNPs 
to illustrate the stratified qq-plots.
These data simulated as described in Chen et al or in Milet et al with the program `ms` 
(Hudson 2003).
We created a data package for these data. It includes 600k SNPs at linkage equilibrium 
for 1000 individuals simulated on a 20x20 grid. It includes some first-order related individuals.
A binary phenotype was simulated with a large population structure effect as described
in Milet et al.

## Loading the data

The package needs to be installed first.

```{r eval=FALSE}
install.packages("GridData", repos="https://genostats.github.io/R/")
```

Then the data are easily loaded with
```{r readGrid}
filepath <-system.file("extdata", "GridData.bed", package="GridData")
x <- read.bed.matrix(filepath)
x <- set.stats(x)
```

A look at the data size:
```{r}
x
```
At the two population strata, which are included in the `x@ped` data frame:
```{r}
table(x@ped$pop)
```
At the simulated phenotype:
```{r}
table(x@ped$pheno)
```
And finally at the relation between strata and phenotype
```{r}
table(x@ped$pop, x@ped$pheno)
```

## Running the association test

The GRM is computed with Gaston, here on the whole set of SNPs (at linkage equilibrium).

```{r GRM, cache=TRUE}
K <- GRM(x)
eigenK <- eigen(K)
```

```{r associationTests, cache = TRUE}
MLM <- association.test(x, method = "lmm", response = "quantitative", test = "wald", eigenK = eigenK, p = 10)
MLR <- association.test.logistic(x,  K = K, eigenK = eigenK, p = 10, algorithm = "amle")
```

Now we can draw stratified quantile-quantile plots. We need to defined SNP categories,
which can be done either with the 'true' strata information (as described in Chen et al), 
or with the first PCs (Milet et al).

## Drawing  stratified QQ-plots

The function `SNP.category` can handle any kind of variable: strata information, or a Principal Component 
as a proxy.

```{r}
# SNPs categories from true strata information
cat.str <- SNP.category(x,x@ped$pop) 
# SNPs categories from the first PCs coordinates
cat.PC1 <- SNP.category(x,-eigenK$vectors[,1]) 
cat.PC2 <- SNP.category(x,-eigenK$vectors[,2])
```

First, the QQ-plots obtained with the mixed linear model, which shows differences between SNP categories:
```{r qqplotsMLM, fig.height = 3, fig.width = 9, dev = 'png', cache = TRUE}
par(mfrow=c(1,3), cex = 0.7)
qqplot.pvalues(MLM, cat.str, col.abline="blue", main="MLM - Variance ratio from strata", pch = 16)
qqplot.pvalues(MLM, cat.PC1, col.abline="blue", main="MLM - Variance ratio from PC1", pch = 16)
qqplot.pvalues(MLM, cat.PC2, col.abline="blue", main="MLM - Variance ratio from PC2", pch = 16)
```

Then, the QQ-plot obtained with mixed logistic regression, in which no such differences exist.
```{r qqplotsMLR, fig.height = 3, fig.width = 9, dev = 'png', cache = TRUE}
par(mfrow=c(1,3), cex = 0.7)
qqplot.pvalues(MLR, cat.str, col.abline="blue", main="MLR - Variance ratio from strata", pch = 16)
qqplot.pvalues(MLR, cat.PC1, col.abline="blue", main="MLR - Variance ratio from PC1", pch = 16)
qqplot.pvalues(MLR, cat.PC1, col.abline="blue", main="MLR - Variance ratio from PC2", pch = 16)
```

# References

* Chen et al.
* Milet and Perdry.
* Hudson.
* Zhou et al.


